---
title: "PS270 Final Project"
author: "Maggie McEllistrem"
description: My final project
output: html_document
---



## Project thoughts

I am interested in exploring data related to the difference presidents influence has on the economy. Each president has different implementations that they make during their term and I think gathering data based on the success of each president during there term would be very interesting to compare. 

```{r}
library(readr)
```



How does each president have a negative or positive impact on the economy during their presidency?Typically a republican president would be assumed to make the economy more successful rather than a democratic president. Republican presidents focus more on businesses, deregulation within businesses, tax reduction and energy affordability. These traits of a typical republican president would likely boost the success of the economy. The explanatory variables are whether there was a republican president in a given year. I will measure their economic success within their presidential terms. The economic success during the term. This will be measured by GDP, unemployment rate, inflation rate, misery index, stock market performance, national debt. Greater economic success during the Reagan and G.W Bush administrations would prove my thesis correct while greater success within the economy during the Clinton, and Carter administrations would prove that my hypothesis is wrong.


```{r}
my_data <- read_delim("gdpv6.txt")
```
```{r}
head(my_data)
```
```{r}
library(ggplot2)
my_data |>
  ggplot(mapping = aes(x = year, y = realgdp)) +
  geom_col(fill = "steelblue") + 
  labs(y = "GDP",
       x = "Presidency Years",
       title = "GDP rate per Presidency")

```

```{r}
gdp_fit <- lm(`realgdp` ~ year, data = my_data)
gdp_fit
```
The graph shows a steady rise in GDP overall. There are drastic rises in the growth of GDP during the Reagan(1981-1989) and G.W Bush(2001-2009) administrations than during the Clinton(1993-2001) or Carter(1977-1981) administrations. Overall from the comparison of these presidencies, republican presidents to the democratic presidents, republicans are more likely to boost the economy than democratic presidents are.


